---
title: "Advanced Regression Methods: Poisson, Negative Binomial, Ordinal, Truncated, and Tobit"
author: "Mintay Misgano, PhD"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

---

## Overview

Standard linear regression assumes a normally distributed, continuous, unbounded dependent variable. When the outcome violates these assumptions, specialized regression models are required. This document demonstrates five such models:

| Model | DV Type | Dataset | Question |
|-------|---------|---------|----------|
| Poisson | Count (non-negative integer) | awards.csv | What predicts the number of academic awards? |
| Negative Binomial | Overdispersed count | awards.csv | Same question when variance > mean |
| Ordinal Logistic | Ordered categories | world_values.csv | What predicts perceived poverty level? |
| Truncated Regression | Continuous, sampled range only | scores.csv | What predicts academic scores in a truncated sample? |
| Tobit | Censored at boundary | AER::Affairs (built-in) | What predicts affair frequency when many report zero? |

---

## Setup

```{r packages}
library(tidyverse)
library(ggplot2)
library(MASS)
library(AER)
library(truncreg)
library(psych)
library(knitr)
```

---

## Part 1: Poisson Regression

### Background

Poisson regression models count data — non-negative integers that represent the number of times an event occurs. The model assumes that the conditional mean equals the conditional variance (equidispersion). Here I predict the number of academic awards (`num_awards`) earned by students based on their programme of study (`prog`), maths score (`math`), age, and gender.

### 1.1 Load and Explore

```{r poisson-load}
awards <- read.csv("GitHub_Ready/01a_awards.csv")
awards$prog <- factor(awards$prog, labels = c("General","Academic","Vocational"))
awards$gender <- factor(awards$gender, labels = c("Female","Male"))
str(awards)
```

```{r poisson-desc}
describe(awards[, c("num_awards","math","age")]) |>
  round(2) |>
  kable(caption = "Table 1.1. Descriptive Statistics — Awards Dataset")
```

```{r poisson-dist, fig.cap="Figure 1.1. Distribution of num_awards by programme"}
ggplot(awards, aes(x = num_awards, fill = prog)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("#616161","#1976D2","#388E3C")) +
  labs(title   = "Distribution of Number of Awards by Programme",
       subtitle = "Count data with right skew — Poisson regression appropriate",
       x = "Number of Awards", y = "Count", fill = "Programme") +
  theme_minimal(base_size = 12)
```

### 1.2 Poisson Model

```{r poisson-model}
poisson_m1 <- glm(num_awards ~ prog + math, family = "poisson", data = awards)
summary(poisson_m1)
```

```{r poisson-irr}
# Incidence Rate Ratios (exponentiated coefficients)
irr <- exp(cbind(IRR = coef(poisson_m1), confint(poisson_m1)))
kable(round(irr, 3), caption = "Table 1.2. Poisson — Incidence Rate Ratios (95% CI)")
```

### 1.3 Overdispersion Test

The key diagnostic for Poisson models: if variance substantially exceeds the mean, the equidispersion assumption is violated and Negative Binomial is preferred.

```{r poisson-dispersion}
dispersiontest(poisson_m1, trafo = 1)
```

```{r poisson-predict, fig.cap="Figure 1.2. Observed vs. predicted award counts (Poisson)"}
awards$pred_poisson <- predict(poisson_m1, type = "response")

ggplot(awards, aes(x = math, y = num_awards, color = prog)) +
  geom_jitter(height = 0.15, alpha = 0.5, size = 1.5) +
  geom_line(aes(y = pred_poisson), linewidth = 1) +
  scale_color_manual(values = c("#616161","#1976D2","#388E3C")) +
  labs(title   = "Poisson Model: Predicted Award Count vs. Math Score",
       x = "Math Score", y = "Number of Awards", color = "Programme") +
  theme_minimal(base_size = 12)
```

---

## Part 2: Negative Binomial Regression

### Background

When the dispersion test indicates overdispersion (variance > mean), Negative Binomial regression relaxes the equidispersion constraint by estimating an additional dispersion parameter (θ). It consistently fits count data better than Poisson when overdispersion is present.

```{r nb-model}
nb_m1 <- glm.nb(num_awards ~ prog + math, data = awards)
summary(nb_m1)
```

```{r nb-irr}
irr_nb <- exp(cbind(IRR = coef(nb_m1), confint(nb_m1)))
kable(round(irr_nb, 3), caption = "Table 2.1. Negative Binomial — Incidence Rate Ratios (95% CI)")
```

### 2.1 Model Comparison: Poisson vs. Negative Binomial

```{r nb-compare}
cat("=== AIC Comparison ===\n")
cat(sprintf("Poisson AIC          : %.2f\n", AIC(poisson_m1)))
cat(sprintf("Negative Binomial AIC: %.2f\n", AIC(nb_m1)))
cat(sprintf("\nNB theta (dispersion): %.3f\n", nb_m1$theta))
cat("Lower theta = more overdispersion; Poisson is the NB limit as theta → ∞\n")

# Likelihood ratio test
lrt <- anova(poisson_m1, nb_m1, test = "Chisq")
print(lrt)
```

```{r nb-extended}
# Extended model adding age and gender
nb_m2 <- glm.nb(num_awards ~ prog + math + age + gender, data = awards)
summary(nb_m2)
cat("\nAIC (extended NB):", AIC(nb_m2), "\n")
```

---

## Part 3: Ordinal Logistic Regression

### Background

Ordinal logistic regression (`polr`) models an ordered categorical DV — where categories have a meaningful rank but unequal intervals. Here I predict perceived government effort to reduce poverty (`poverty`: Too Little / About Right / Too Much) from religion, education, country, age, and gender in a World Values Survey dataset (N = 5,381).

### 3.1 Load and Explore

```{r ordinal-load}
wv <- read.csv("GitHub_Ready/01b_world_values.csv")
wv$poverty  <- factor(wv$poverty,  levels = c("Too Little","About Right","Too Much"), ordered = TRUE)
wv$religion <- factor(wv$religion)
wv$degree   <- factor(wv$degree)
wv$country  <- factor(wv$country)
wv$gender   <- factor(wv$gender)
str(wv)
```

```{r ordinal-desc}
cat("Poverty level distribution:\n")
print(prop.table(table(wv$poverty)) * 100 |> round(1))
```

```{r ordinal-bar, fig.cap="Figure 3.1. Poverty perception by country and gender"}
ggplot(wv, aes(x = country, fill = poverty)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#1976D2","#FFA726","#C62828")) +
  facet_wrap(~gender) +
  labs(title   = "Perceived Poverty Reduction Effort by Country and Gender",
       subtitle = "Proportion in each ordered response category",
       x = "Country", y = "Proportion", fill = "Poverty Perception") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 3.2 Ordinal Model

```{r ordinal-model}
ord_m1 <- polr(poverty ~ religion + degree + country + age + gender,
               data = wv, Hess = TRUE)
summary(ord_m1)
```

```{r ordinal-pvalues}
# Compute p-values (polr doesn't include them by default)
results <- coef(summary(ord_m1))
p_vals  <- pnorm(abs(results[, "t value"]), lower.tail = FALSE) * 2
results <- cbind(results, `p value` = round(p_vals, 4))
kable(round(results, 4), caption = "Table 3.1. Ordinal Model Coefficients with p-values")
```

```{r ordinal-or}
# Odds Ratios
or_table <- exp(cbind(OR = coef(ord_m1), confint(ord_m1)))
kable(round(or_table, 3), caption = "Table 3.2. Ordinal Model — Odds Ratios (95% CI)")
```

```{r ordinal-predict}
# Predicted probabilities for a representative profile
new_wv <- data.frame(religion = "yes", degree = "yes", country = "USA",
                     age = 40, gender = "female")
pred_probs <- predict(ord_m1, newdata = new_wv, type = "p")
cat("Predicted probabilities for: religious, college-educated, 40-year-old US female\n")
round(pred_probs, 3)
```

---

## Part 4: Truncated Regression

### Background

Truncated regression addresses samples where observations below (or above) a certain threshold are systematically excluded — not just censored. Here the dataset contains academic scores (`score`) for students, but only those whose scores fall within an observed range (scores below 40 are absent; very high scores above 100 are also missing). OLS on a truncated sample produces biased estimates; `truncreg` corrects this.

### 4.1 Load and Explore

```{r truncated-load}
scores_df <- read.csv("GitHub_Ready/01c_scores.csv")
scores_df$prog <- factor(scores_df$prog)
str(scores_df)
```

```{r truncated-hist, fig.cap="Figure 4.1. Distribution of scores — evidence of truncation at ceiling"}
ggplot(scores_df, aes(x = score, fill = prog)) +
  geom_histogram(binwidth = 3, alpha = 0.8, position = "identity") +
  scale_fill_manual(values = c("#1976D2","#388E3C","#F57C00")) +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 97, y = 12, label = "Ceiling\ntruncation", color = "red",
           size = 3, hjust = 1) +
  labs(title   = "Score Distribution by Programme",
       subtitle = "Right truncation at score = 100 (no observations beyond ceiling)",
       x = "Academic Score", y = "Count", fill = "Programme") +
  theme_minimal(base_size = 12)
```

### 4.2 Truncated Regression Models

```{r truncated-model}
# Right-truncated at 100 (scores above 100 unobservable)
trunc_right <- truncreg(score ~ langscore + prog, data = scores_df,
                        point = 100, direction = "right")
summary(trunc_right)
```

```{r truncated-model2}
# Left-truncated at 60 (students below 60 may have dropped/not recorded)
trunc_left <- truncreg(score ~ langscore + prog, data = scores_df,
                       point = 60, direction = "left")
summary(trunc_left)
```

```{r truncated-compare}
cat("=== OLS vs. Truncated Regression — Coefficient Comparison ===\n")
ols_m <- lm(score ~ langscore + prog, data = scores_df)
ols_coefs   <- coef(ols_m)[1:2]
trunc_coefs <- coef(trunc_right)[1:2]

comparison <- data.frame(
  Parameter  = c("Intercept","langscore"),
  OLS        = round(ols_coefs,   3),
  Truncated  = round(trunc_coefs, 3),
  Difference = round(trunc_coefs - ols_coefs, 3)
)
kable(comparison, caption = "Table 4.1. OLS vs. Truncated Regression Coefficient Comparison")
```

---

## Part 5: Tobit Regression (Censored Regression)

### Background

Tobit regression addresses censored outcomes — where the DV is observed only within a bounded range, with values piled up at one or both boundaries. Unlike truncation (where out-of-range observations are absent), censoring means we observe the individual but only know their value is at or beyond the boundary. The classic example is `affairs` from the `AER::Affairs` dataset: many respondents report exactly zero affairs (the floor), but the true propensity to have affairs varies continuously.

### 5.1 Load and Explore

```{r tobit-load}
data("Affairs", package = "AER")
affairs_df <- Affairs
str(affairs_df)
```

```{r tobit-desc}
cat("Affairs distribution:\n")
print(table(affairs_df$affairs))
cat(sprintf("\nZero-inflated: %.1f%% of respondents report 0 affairs\n",
            mean(affairs_df$affairs == 0) * 100))
```

```{r tobit-hist, fig.cap="Figure 5.1. Distribution of affairs — heavy mass at zero (censored from below)"}
ggplot(affairs_df, aes(x = affairs)) +
  geom_histogram(binwidth = 1, fill = "#7B1FA2", alpha = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 400, label = "Censoring\npoint (0)",
           color = "red", size = 3.5, hjust = 0) +
  labs(title   = "Distribution of Extramarital Affairs",
       subtitle = "Heavy censoring at zero — Tobit model appropriate",
       x = "Number of Affairs", y = "Count") +
  theme_minimal(base_size = 12)
```

### 5.2 Tobit Models

```{r tobit-model}
# Two-limit Tobit: censored both below 0 and above 4 (max per year in data)
tobit_m1 <- tobit(affairs ~ age + yearsmarried + religiousness + occupation + rating,
                  left = 0, right = 4, data = affairs_df)
summary(tobit_m1)
```

```{r tobit-model2}
# Left-censored only at 0
tobit_m2 <- tobit(affairs ~ age + yearsmarried + religiousness + occupation + rating,
                  left = 0, data = affairs_df)
summary(tobit_m2)
```

```{r tobit-compare}
cat("=== OLS vs. Tobit — Coefficient Comparison ===\n")
ols_aff   <- lm(affairs ~ age + yearsmarried + religiousness + occupation + rating,
                data = affairs_df)
ols_c     <- coef(ols_aff)
tobit_c   <- coef(tobit_m2)[1:6]

comp_aff <- data.frame(
  Predictor  = names(ols_c),
  OLS        = round(ols_c,   3),
  Tobit      = round(tobit_c, 3),
  Difference = round(tobit_c - ols_c, 3)
)
kable(comp_aff, caption = "Table 5.1. OLS vs. Tobit Coefficient Comparison (left-censored at 0)")
```

```{r tobit-margins, fig.cap="Figure 5.2. Tobit predicted probability of any affair by marriage rating"}
pred_range <- data.frame(
  age          = mean(affairs_df$age),
  yearsmarried = mean(affairs_df$yearsmarried),
  religiousness = mean(affairs_df$religiousness),
  occupation   = mean(affairs_df$occupation),
  rating       = 1:5
)
pred_range$pred <- predict(tobit_m2, newdata = pred_range, type = "response")

ggplot(pred_range, aes(x = rating, y = pred)) +
  geom_line(color = "#7B1FA2", linewidth = 1) +
  geom_point(size = 4, color = "#7B1FA2") +
  labs(title   = "Tobit: Predicted Affairs by Marriage Rating",
       subtitle = "All other predictors held at sample means",
       x = "Marriage Happiness Rating (1 = very unhappy, 5 = very happy)",
       y = "Predicted Number of Affairs") +
  theme_minimal(base_size = 12)
```

---

## Summary: When to Use Each Model

```{r summary-table}
model_summary <- data.frame(
  Model = c("Poisson","Negative Binomial","Ordinal Logistic","Truncated","Tobit"),
  `DV Structure` = c(
    "Count (integer ≥ 0)",
    "Overdispersed count (variance > mean)",
    "Ordered categories (rank matters, intervals unequal)",
    "Continuous — out-of-range obs. entirely absent",
    "Continuous — obs. present but value capped/floored"
  ),
  `Key Assumption` = c(
    "E(Y) = Var(Y) (equidispersion)",
    "Var(Y) > E(Y) (adds dispersion param θ)",
    "Proportional odds across categories",
    "Truncation point and direction specified correctly",
    "Censoring boundary and direction specified correctly"
  ),
  `Example Use Case` = c(
    "Absence days, incident counts, event frequencies",
    "Same — when counts are overdispersed",
    "Ordered survey ratings or ranked judgments",
    "Samples with observations excluded beyond a threshold",
    "Ratings or outcomes censored at an upper or lower boundary"
  )
)

kable(model_summary,
      col.names = c("Model","DV Structure","Key Assumption","Example Use Case"),
      caption   = "Table 6.1. Regression Model Selection Guide")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
